---
title: "Supplementary tables for carotovoricin analysis"
author: "Lakhansing Pardeshi"
date: "`r Sys.Date()`"
format: 
  html: 
    embed-resources: true
    df-print: paged
knitr: 
  opts_chunk:
    fig.height: 7
---


```{r}
#| label: setup
#| echo: false
#| warning: false

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(org.Pectobacterium.spp.pan.eg.db))
suppressPackageStartupMessages(library(DT))
suppressPackageStartupMessages(library(openxlsx))

rm(list = ls())

source("https://raw.githubusercontent.com/lakhanp1/omics_utils/main/RScripts/utils.R")
source("scripts/utils/config_functions.R")
source("scripts/utils/homology_groups.R")

################################################################################
set.seed(124)

confs <- prefix_config_paths(
  conf = suppressWarnings(configr::read.config(file = "project_config.yaml")),
  dir = "."
)

pangenome <- confs$data$pangenomes$pectobacterium.v2$name
panConf <- confs$data$pangenomes[[pangenome]]

panOrgDb <- org.Pectobacterium.spp.pan.eg.db

```

```{r}
#| echo: false
#| warning: false

## write tables to excel file
wb <- openxlsx::createWorkbook()

add_table_to_workbook <- function(data, wb, table_num, description = "## Table"){
  currentSheet <- paste("Table_S", table_num, sep = "")
  openxlsx::addWorksheet(wb, sheetName = currentSheet)
  
  openxlsx::writeData(
    wb = wb, sheet = currentSheet,
    x = description
  )
  openxlsx::writeDataTable(
    wb = wb, x = data, sheet = currentSheet,
    startCol = 1, startRow = 2, withFilter = TRUE,
    keepNA = TRUE, na.string = "NA"
  )
  openxlsx::freezePane(wb = wb, sheet = currentSheet, firstActiveRow = 3, firstActiveCol = 2)
  
  return(wb)
}

table_counter <- 1
```

::: {.panel-tabset}

## Table S`r table_counter`

Metadata for all the genome in the current pangenome along with carotovoricin presence/absence information

```{r}
#| echo: false
#| column: screen-inset-right

metadata <- suppressMessages(readr::read_csv(panConf$files$metadata)) %>% 
  dplyr::select(-starts_with("nodepath."), -Genome) %>% 
  dplyr::select(
    genomeId, SpeciesName, sampleId, strain, AssemblyAccession, geo_loc_country, collection_year,
    host, env_broad_scale, type_material, AssemblyName, BioprojectAccn,
    BioSampleAccn, length, N50, L50, n_contigs
  )

wb <- add_table_to_workbook(
  data = metadata, wb = wb, table_num = table_counter,
  description = paste(
    "## Table S", table_counter, ": Metadata for genome assemblies used in the current pangenome",
    sep = ""
  )
)

DT::datatable(
  data = metadata,
  rownames = FALSE,
  filter = "top",
  class = 'compact hover',
  extensions = c('KeyTable', 'Scroller', 'Select', 'SearchBuilder', 'FixedColumns'),
  options = list(
    autoWidth = FALSE,
    dom = 'Qlfrtip',
    scrollX = TRUE,
    fixedColumns = list(leftColumns = 2),
    keys = TRUE,
    scroller = TRUE,
    scrollY = 600
  ),
  selection = 'none'
)

table_counter <- table_counter + 1
```

## Table S`r table_counter`

geNomad prophage prediction result summary for 454 genomes

```{r}
#| echo: false
#| column: screen-inset-right

prophages <- suppressMessages(readr::read_tsv(confs$data$prophages$files$data)) %>% 
  dplyr::mutate(
    integrated = dplyr::if_else(
      condition = is.na(start) & is.na(end), true = "N", false = "Y" 
    )
  ) %>% 
  dplyr::select(prophage_id, genomeId, everything()) %>% 
  dplyr::relocate(integrated, .after = end) %>% 
  dplyr::left_join(
    y = dplyr::select(metadata, genomeId, strain),
    by = "genomeId"
  ) %>%
  dplyr::relocate(strain, .after = SpeciesName)

wb <- add_table_to_workbook(
  data = prophages, wb = wb, table_num = table_counter,
  description = paste(
    "## Table S", table_counter, ": geNomad prophage prediction result summary for 454 genomes",
    sep = ""
  )
)

DT::datatable(
  data = prophages,
  rownames = FALSE,
  filter = "top",
  class = 'compact hover',
  extensions = c('KeyTable', 'Scroller', 'Select', 'SearchBuilder', 'FixedColumns'),
  options = list(
    autoWidth = FALSE,
    dom = 'Qlfrtip',
    scrollX = TRUE,
    fixedColumns = list(leftColumns = 2),
    keys = TRUE,
    scroller = TRUE,
    scrollY = 600
  ),
  selection = 'none'
)

table_counter <- table_counter + 1

```


## Table S`r table_counter`

Prophage clustering based on syntenic Jaccard index

```{r}
#| echo: false
#| column: screen-inset-right

phage_grps <- suppressMessages(
  readr::read_tsv(confs$analysis$prophages$files$clusters)
) %>% 
  dplyr::select(-starts_with("nodepath.")) %>%
  dplyr::left_join(
    y = dplyr::select(metadata, genomeId, strain),
    by = "genomeId"
  ) %>%
  dplyr::relocate(strain, .after = SpeciesName)

wb <- add_table_to_workbook(
  data = phage_grps, wb = wb, table_num = table_counter,
  description = paste(
    "## Table S", table_counter, ": Prophage clustering based on syntenic Jaccard index",
    sep = ""
  )
)

DT::datatable(
  data = phage_grps,
  rownames = FALSE,
  filter = "top",
  class = 'compact hover',
  extensions = c('KeyTable', 'Scroller', 'Select', 'SearchBuilder', 'FixedColumns'),
  options = list(
    autoWidth = FALSE,
    dom = 'Qlfrtip',
    scrollX = TRUE,
    fixedColumns = list(leftColumns = 2),
    keys = TRUE,
    scroller = TRUE,
    scrollY = 600
  ),
  selection = 'none'
)

table_counter <- table_counter + 1

```

## Table S`r table_counter`

Carotovoricin cluster presence/absence data

```{r}
#| echo: false
#| column: screen-inset-right

ctvPav <- suppressMessages(
  readr::read_tsv(file = confs$analysis$ctv$data$files$ctv_pav)
) |> 
  dplyr::rename(
    "n_homology_groups" = nHgs,
    "homology_group_signature" = hgs,
    "coordinates" = pos
  )

wb <- add_table_to_workbook(
  data = ctvPav, wb = wb, table_num = table_counter,
  description = paste(
    "## Table S", table_counter, ": Carotovoricin presence.absence data for 454 genomes.",
    "In case of carotovoricin cluster split onto multiple contigs, \"|\" seperator is used for coordinates and homology_group_signature columns",
    sep = ""
  )
)

DT::datatable(
  data = ctvPav,
  rownames = FALSE,
  filter = "top",
  class = 'compact hover',
  extensions = c('KeyTable', 'Scroller', 'Select', 'SearchBuilder', 'FixedColumns'),
  options = list(
    autoWidth = FALSE,
    dom = 'Qlfrtip',
    scrollX = TRUE,
    fixedColumns = list(leftColumns = 2),
    keys = TRUE,
    scroller = TRUE,
    scrollY = 600
  ),
  selection = 'none'
)

table_counter <- table_counter + 1
```

## Table S`r table_counter`

Carotovoricin cluster homology groups and their functional categories

```{r}
#| echo: false
#| column: screen-inset-right

ctvHgs <- suppressMessages(
  readr::read_tsv(file = confs$analysis$ctv$data$files$hg_broad_functions)
) |> 
  dplyr::rename(homology_group = hgId)

wb <- add_table_to_workbook(
  data = ctvHgs, wb = wb, table_num = table_counter,
  description = paste(
    "## Table S", table_counter, ": Carotovoricin cluster homology groups and their functional categories",
    sep = ""
  )
)

DT::datatable(
  data = ctvHgs,
  rownames = FALSE,
  filter = "top",
  class = 'compact hover',
  extensions = c('KeyTable', 'Scroller', 'Select', 'SearchBuilder', 'FixedColumns'),
  options = list(
    autoWidth = FALSE,
    dom = 'Qlfrtip',
    scrollX = TRUE,
    fixedColumns = list(leftColumns = 3),
    keys = TRUE,
    scroller = TRUE,
    scrollY = 600
  ),
  selection = 'none'
)

table_counter <- table_counter + 1

```

## Table S`r table_counter`

Homology group based haplotypes for the tail fiber locus

```{r}
#| echo: false
#| column: screen-inset-right

tfl_haplotypes <-  suppressMessages(
  readr::read_tsv(file = confs$analysis$ctv$tfl$files$hg_regions)
) |> 
  dplyr::add_count(haplotype, name = "n") |> 
  dplyr::rename(
    "n_homology_groups" = nHgs,
    "homology_group_signature" = hgs
  )

wb <- add_table_to_workbook(
  data = tfl_haplotypes, wb = wb, table_num = table_counter,
  description = paste(
    "## Table S", table_counter, ": Carotovoricin tail fiber locus haplotypes",
    sep = ""
  )
)

DT::datatable(
  data = tfl_haplotypes,
  rownames = FALSE,
  filter = "top",
  class = 'compact hover',
  extensions = c('KeyTable', 'Scroller', 'Select', 'SearchBuilder', 'FixedColumns'),
  options = list(
    autoWidth = FALSE,
    dom = 'Qlfrtip',
    scrollX = TRUE,
    fixedColumns = list(leftColumns = 3),
    keys = TRUE,
    scroller = TRUE,
    scrollY = 600
  ),
  selection = 'none'
)

table_counter <- table_counter + 1
```


:::

```{r}
#| echo: false

# save all table to Excel file
openxlsx::saveWorkbook(
  wb = wb, overwrite = TRUE,
  file = file.path("reports/manuscript_ctv", "ms_ctv_supplementary_data.xlsx")
)

```

