---
title: "CTV summary across *Pectobacterium* genus"
date: "`r Sys.time()`"
knitr: 
  opts_chunk:
    fig.height: 7
---

This script will summarize CTV in the *Pectobacterium* genus, including the following steps:

- Intact CTV statistics
- Haplotype summary for various CTV regions

## Initial setup

```{r}
#| label: setup
#| warning: false

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ape))
suppressPackageStartupMessages(library(treeio))
suppressPackageStartupMessages(library(ggtree))
suppressPackageStartupMessages(library(ggbeeswarm))
suppressPackageStartupMessages(library(ggrastr))
suppressPackageStartupMessages(library(cowplot))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(org.Pectobacterium.spp.pan.eg.db))

rm(list = ls())

source("https://raw.githubusercontent.com/lakhanp1/omics_utils/main/RScripts/utils.R")
source("scripts/utils/config_functions.R")
source("scripts/utils/phylogeny_functions.R")
source("scripts/utils/homology_groups.R")
source("scripts/utils/heatmap_utils.R")
################################################################################
set.seed(124)

confs <- prefix_config_paths(
  conf = suppressWarnings(configr::read.config(file = "project_config.yaml")),
  dir = "."
)

outDir <- paste(confs$analysis$ctv$ctv_region$path)

pangenome <- confs$data$pangenomes$pectobacterium.v2$name
panConf <- confs$data$pangenomes[[pangenome]]
panOrgDb <- org.Pectobacterium.spp.pan.eg.db
prophageLenCutoff <- confs$analysis$prophages$cutoff_length
treeMethod <- "core_snp_ml" # ani_upgma, kmer_nj

```

## Import data

```{r}
sampleInfo <- get_metadata(file = panConf$files$metadata, genus = confs$genus)

sampleInfoList <- as.list_metadata(
  df = sampleInfo, sampleId, sampleName, SpeciesName, strain, nodeLabs, genomeId
)

rawTree <- import_tree(
  confs$analysis$phylogeny[[treeMethod]]$files$tree_rooted
)

# rawTree <- ape::ladderize(rawTree, right = FALSE)

speciesOrder <- suppressMessages(
  readr::read_tsv(confs$analysis$phylogeny[[treeMethod]]$files$species_order)
) %>%
  dplyr::mutate(order = n():1)


```

## Carotovoricin presence-absence across the *Pectobacterium* genus

```{r}

ctvPav <- suppressMessages(
  readr::read_tsv(file = confs$analysis$ctv$data$files$ctv_pav)
) |> 
  dplyr::left_join(
    y = dplyr::select(
      sampleInfo, genomeId, strain, AssemblyAccession, geo_loc_country,
      collection_year, isolation_source
      ),
    by = "genomeId"
  )

intact_ctv_genomes <- dplyr::filter(ctvPav, ctv_presence == "present") |> 
  dplyr::pull(genomeId)

# create a descriptive lable for phylogenetic tree
label_cols <- c("SpeciesName", "collection_year", "geo_loc_country", "AssemblyAccession")
names(label_cols) <- paste("padded_", label_cols, sep = "")

ctvPav <- dplyr::mutate(
    ctvPav,
    dplyr::across(
      .cols =  tidyselect::all_of(label_cols),
      .fns = ~stringi::stri_sprintf(
        format = paste("%-", max(stringr::str_length(.x), na.rm = TRUE) + 1, "s", sep = ""),
        .x,
        na_string = ""
      )
    )
  ) |> 
    tidyr::unite(col = "nodeLabs", tidyselect::all_of(names(label_cols)), sep = " ") |> 
    dplyr::select(genomeId, everything())

```

Carotovoricin across 454 *Pectobacterium* genomes:

```{r}
table(ctvPav$ctv_presence)
```

## *Pectobacterium* phylogeny with carotovoricin metadata

```{r}
legend_size <- 18
legend_text_size <- 22
outGroup <- confs$analysis$phylogeny$outgroup

treeTbl <- tidytree::as_tibble(rawTree) %>%
  dplyr::full_join(y = sampleInfo, by = c("label" = "genomeId")) %>%
  treeio::as.treedata()

pt_tree <- ggtree::ggtree(treeTbl) +
  geom_treescale(
    x = 0, y = nrow(sampleInfo) * 0.8,
    fontsize = 8, linesize = 2
  )

longestBranch <- which(pt_tree$data$x > confs$analysis$phylogeny[[treeMethod]]$cut_branch)
pt_tree$data$x[longestBranch] <- confs$analysis$phylogeny[[treeMethod]]$cut_branch

rootNode <- rootnode(treeTbl)

pt_phy_theme <- ggplot2::theme_bw() +
  theme(
    axis.text.y = element_blank(),
    # panel.grid = element_blank(),
    panel.background = element_blank(),
    # panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.text = element_text(size = legend_text_size),
    legend.title = element_text(size = legend_text_size + 2, face = "bold"),
    axis.title = element_blank()
  )

# bootstrap values
pt_tree <- pt_tree +
  ggtree::geom_point2(
    mapping = aes(
      subset = !isTip & node != rootNode, 
      color = cut(bootstrap, c(0, 50, 70, 90, 100))),
    shape = 16, size = 2
  ) +
  scale_color_manual(
    values = c("red", "#f781bf", "blue", "black"),
    guide = guide_legend(override.aes = list(size = legend_size), order = 1),
    name = 'Bootstrap %', 
    breaks = c('(0,50]', '(50,70]', '(70,90]'), 
    labels = expression(BP < "50%", "50%" <= BP * " < 70%", "70%" <= BP * " < 90%"),
    na.value = "transparent"
  ) +
  ggnewscale::new_scale_color()

## mark outgroup
pt_tree2 <- mark_outgroup(
  pt = pt_tree, otg = outGroup, column = "sampleId"
) + 
  theme(
    legend.text = element_text(size = legend_text_size),
    legend.title = element_text(size = legend_text_size + 2, face = "bold")
  )

pt_tree3 <- pt_tree2 +
  # ggtree::geom_nodelab(
  #   mapping = aes(label = label),
  #   node = "internal", size = 3, hjust = 1.3, color = "red"
  # ) +
  ggtree::geom_tiplab(
    mapping = aes(label = label, color = type_material),
    size = 3, align = TRUE, linesize = 0.5
  ) +
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.05))) +
  scale_color_manual(
    values = c("type strain" = "blue"),
    name = "Type strain",
    guide = guide_legend(
      order = 2, 
      override.aes = list(size = legend_size)
    ),
    na.value = "black"
  ) +
  ggnewscale::new_scale_color()

## add metadata
pt_labels <- dplyr::select(ctvPav, label = genomeId, nodeLabs) |>
  ggplot2::ggplot(mapping = aes(x = "desc", y = label)) +
  ggplot2::geom_text(
    mapping = aes(label = nodeLabs),
    hjust = 0, family = "mono"
  ) +
  ggplot2::scale_x_discrete(
    expand = expansion(mult = c(0.01, 0.5)), position = "top"
  ) +
  pt_phy_theme +
  theme(axis.text.x = element_blank())

pt_ctv_pav <- dplyr::select(ctvPav, label = genomeId, ctv_presence) |>
  ggplot2::ggplot(mapping = aes(x = "ctv", y = label)) +
  ggplot2::geom_text(
    mapping = aes(label = ctv_presence),
    hjust = 0, family = "mono"
  ) +
  ggplot2::scale_x_discrete(
    expand = expansion(mult = c(0.01, 0.5)), position = "top"
  ) +
  pt_phy_theme +
  theme(axis.text.x = element_blank())

pt_tree4 <- pt_labels |> 
  aplot::insert_left(pt_tree3, width = 1.6) |> 
  aplot::insert_right(pt_ctv_pav, width = 0.3)

ggsave(
  plot = pt_tree4, width = 12, height = 22, scale = 2,
  filename = paste(outDir, "/ctv_detailed_phylo.pdf", sep = "")
)
```

:::{.scrolling_y}

```{r}
#| column: page-right
#| fig-height: 50
#| fig-width: 22
#| out-width: '100%'
#| layout-valign: top
#| echo: false

pt_tree4
```

:::

## Combine and summarize all variable region haplotypes

### Phylogenetic tree for genomes that have intact carotovoricin cluster

```{r}
pt_phy <- ggtree_with_species(
  phy = rawTree@phylo, metadata = sampleInfo,
  genomes = c(intact_ctv_genomes, confs$analysis$phylogeny$outgroup_genome),
  trim_branch = confs$analysis$phylogeny[[treeMethod]]$cut_branch
)
```


```{r}

gg_haplotype_pav <- function(data, fill_color = "black") {
  stopifnot(
    is.data.frame(data),
    all(c("haplotype", "label") %in% colnames(data))
  )
  
  pt_pav <- dplyr::filter(data, !is.na(haplotype) | haplotype != "") |> 
    dplyr::mutate(hap = "y") |> 
    ggplot2::ggplot(data, mapping = aes(x = haplotype, y = label)) +
    geom_tile(mapping = aes(fill = hap), width = 0.9) +
    scale_fill_manual(
      name = NULL,
      values = fill_color,
      na.value = alpha("white", 1),
      guide = NULL,
      na.translate = FALSE
    ) +
    theme_bw() +
    theme(
      panel.grid.major.y = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank(),
      legend.title = element_text(size = 18, face = "bold")
    )
  
  return(pt_pav)
}

```

### Nitric oxide haplotypes

```{r}
nox_data <- suppressMessages(
  readr::read_tsv(file = confs$analysis$ctv$nitric_oxide$files$hg_regions)
) |> 
  dplyr::add_count(haplotype, name = "n")

pt_nox <- gg_haplotype_pav(
  data = dplyr::select(nox_data, label = genomeId, haplotype),
  fill_color = "#30123B"
)

```

### Tail tape measure gene haplotypes

```{r}
tape_data <-  suppressMessages(
  readr::read_tsv(file = confs$analysis$ctv$tape_measure$files$hg_regions)
) |> 
  dplyr::add_count(haplotype, name = "n")

pt_tape <- gg_haplotype_pav(
  data = dplyr::select(tape_data, label = genomeId, haplotype),
  fill_color = "#EF5B12"
)

```

### Tail-fiber locus haplotypes

```{r}
tfl_data <-  suppressMessages(
  readr::read_tsv(file = confs$analysis$ctv$tfl$files$hg_regions)
) |> 
  dplyr::add_count(haplotype, name = "n")

pt_tfl <- gg_haplotype_pav(
  data = dplyr::select(tfl_data, label = genomeId, haplotype, n) |> 
    dplyr::filter(n > 1),
  fill_color = "#FDA532"
)

```

### DNA invertase *ein* (hg_22431685) precence-absence information

```{r}
ein_data <- AnnotationDbi::select(
  x = panOrgDb, keys = "hg_22431685", columns = c("genomeId"), keytype = "GID"
) |> 
  dplyr::mutate(hap = "y") |> 
  dplyr::filter(genomeId %in% .env$intact_ctv_genomes) |> 
  dplyr::mutate(haplotype = "ein")

pt_ein <- gg_haplotype_pav(
  data = dplyr::select(ein_data, label = genomeId, haplotype, hap),
  fill_color = "#46F783"
)

```

```{r}
#| fig-height: 8
#| fig-width: 10
#| out-width: '100%'
#| layout-valign: top
#| echo: false
#| column: page 

pt_haplotypes <- pt_phy$species_key |> 
  aplot::insert_left(pt_phy$tree, width = 3) |> 
  aplot::insert_right(pt_nox, width = 0.2) |> 
  aplot::insert_right(pt_tape, width = 0.5) |> 
  aplot::insert_right(pt_tfl, width = 4) |> 
  aplot::insert_right(pt_ein, width = 0.1)

pdf(file = paste(outDir, "/ctv_haplotypes_phylo.pdf", sep = ""), width = 12, height = 10)
pt_haplotypes
dev.off()

pt_haplotypes
```


